---
title: "Climush_iNat_Manuscript"
author: "Andrew Ratz"
date: "`r Sys.Date()`"
output: html_document
---

`
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggplot2)
library(vegan)

setwd("C:/Users/ratza/OneDrive/Desktop/Kennedy Lab Work/CLIMUSH")

```

```{r Creating spore genus table}

spore.dat <- read.csv("dat.csv")
macro.data.long <- read.csv("macro_long.csv")

#some formatting to fit iNaturalist data
spore.dat <- spore.dat%>%
  mutate(year = year + 2000)%>%
  mutate(date = paste(year, month, day, sep = "-"))

spore.dat$date <- as.Date(spore.dat$date, format = "%Y-%m-%d")

spore.dat$DatePickedUp <- paste("20",spore.dat$DatePickedUp, sep = "")
spore.dat$DateOut <- paste("20",spore.dat$DateOut, sep = "")

spore.dat$DatePickedUp <- as.Date(spore.dat$DatePickedUp, format = "%Y-%m-%d")
spore.dat$DateOut <- as.Date(spore.dat$DateOut, format = "%Y-%m-%d")


temp <- spore.dat%>% filter(is.na(DatePickedUp))%>%
  mutate(DatePickedUp = DateOut + 31)%>%
  mutate(Nr.days.out = as.numeric(DatePickedUp - DateOut))%>%
  mutate(log.days.out = log(Nr.days.out))

spore.temp <- rbind(spore.dat, temp)

spore.dat <- spore.temp%>% filter(!is.na(DatePickedUp))

macro.data.long%>% filter(is.na(ecoregion))

#adjusted spore OTU table
spore_species <- macro.data.long%>%
  filter(!str_detect(Genus, "_sedis"), !is.na(Genus), Species_support >= 0.8)%>%
  group_by(Sample_ID, Species)%>%
  summarise(Count = n())%>%
  left_join(spore.dat[,c(2,9,19,51)], by = "Sample_ID", relationship = "many-to-many", 
            multiple = 'any')%>%
  filter(!is.na(ecoregion), !is.na(season))%>%
  mutate(season = case_when(season == "early" ~ "E", season == "late" ~ "L"))%>%
  mutate(ID = paste(ecoregion,year,season,"S", sep = "_"))

```
I now have a list of all of the species found in the spore samples at each sample with support above 0.8. I now have to pull in the iNat data, filter it to the time frames of the spore traps, and filter the species found within the spores


```{r Loading iNat data}
#the total iNat data includes the transects, so I can load it all in at once
FL_iNat <- read.csv("FL_iNat_30km.csv")
  FL_iNat$common_name <- NULL
  FL_iNat$species_guess <- NULL
  FL_iNat$iconic_taxon_name <- NULL
  FL_iNat$ecoregion <- "SE"
  FL_iNat$site<- "ORD"
  FL_iNat$state<- "FL"
AZ_iNat <- read.csv("AZ_iNat_30km.csv")
  AZ_iNat$ecoregion <- "DSW"
  AZ_iNat$site<- "SRE"
  AZ_iNat$state<- "AZ"
AK_iNat <- read.csv("AK_iNat_30km.csv")
  AK_iNat$ecoregion <- "TAIG"
  AK_iNat$site<- "BNZ"
  AK_iNat$state<- "AK"
CO_iNat <- read.csv("CO_iNat_30km.csv")
  CO_iNat$ecoregion <- "SROC"
  CO_iNat$site<- "NWT"
  CO_iNat$state<- "CO"
KS_iNat <- read.csv("KS_iNat_30km.csv")
  KS_iNat$ecoregion <- "PRAI"
  KS_iNat$site<- "KON"
  KS_iNat$state<- "KS"
MA_iNat <- read.csv("MA_iNat_30km.csv")
  MA_iNat$ecoregion <- "NE"
  MA_iNat$site<- "HARV"
  MA_iNat$state<- "MA"
MN_iNat <- read.csv("MN_iNat_30km.csv")
  MN_iNat$ecoregion <- "GRTL"
  MN_iNat$site<- "CDR"
  MN_iNat$positional_accuracy <- 1
  MN_iNat$positioning_method <- NA
  MN_iNat$state<- "MN"
OR_PIS_iNat <- read.csv("OR_PIS_iNat_30km.csv")
  OR_PIS_iNat$ecoregion <- "PNW"
  OR_PIS_iNat$site<- "PIS"
  OR_PIS_iNat$state<- "OR"
OR_HJA_iNat <- read.csv("OR_HJA_iNat_30km.csv")
  OR_HJA_iNat$ecoregion <- "PNW"
  OR_HJA_iNat$site<- "HJA"
  OR_HJA_iNat$state<- "OR"

iNat_df <- rbind(FL_iNat,AZ_iNat,AK_iNat,CO_iNat,KS_iNat,MA_iNat,MN_iNat,
              OR_PIS_iNat,OR_HJA_iNat)

#some cleaning after the fact
rm(FL_iNat,AZ_iNat,AK_iNat,CO_iNat,KS_iNat,MA_iNat,MN_iNat,
              OR_PIS_iNat,OR_HJA_iNat)

```


```{r Creating transect genus table}
#new df into the pipeline, seeing if this works
transect.dat.raw <- iNat_df %>% dplyr::select(-ecoregion,-site,-state)
transect.dat.raw %>% distinct(taxon_genus_name)
#changing format of date and time
transect.dat <- transect.dat.raw%>%
separate_wider_delim(time_observed_at, " ", names = c("date","time","timezone"),
                       too_few = "align_start")%>%
filter(!is.na(timezone))%>%
mutate(timezone = NULL)

transect.dat$date <- as.Date(transect.dat$date, format = "%Y-%m-%d")#turning into true date format for calculations


transect.dat$year<-str_extract(transect.dat$date,"^[^-]+") #pulling year from the date column

transect.dat <- transect.dat%>%
  mutate(state = case_when(
    place_state_name == "Alaska" ~ "AK", place_state_name == "Arizona" ~ "AZ",
    place_state_name == "Colorado" ~ "CO", place_state_name == "Florida" ~ "FL", 
    place_state_name == "Kansas" ~ "KS", place_state_name == "Oregon" ~ "OR",
    place_state_name == "Massachusetts" ~ "MA", place_state_name == "Minnesota" ~ "MN"))


#When were spores collected
library(lubridate)

spore.dat <- spore.dat%>%
  mutate(start_date = date - days(14), end_date = DatePickedUp)%>%
  arrange(state, year, start_date)
  
samp.timeframe <- spore.dat%>%  
  group_by(state, year)%>%
  mutate(
    time_gap = as.numeric(start_date - lag(end_date, default = first(start_date))),
    group = cumsum(time_gap > 14))%>% #2 weeks based on earlier studies
  group_by(state, year, group)%>%
  summarise(start_date = min(start_date), 
            end_date = max(end_date), .groups = "drop")%>%
  dplyr::select(-group)


library(fuzzyjoin)
#filtering data only found in sampling timeframe I want
transect.dat <- fuzzy_inner_join(
  transect.dat, samp.timeframe,
  by = c("state" = "state", "year" = "year", "date" = "start_date", "date" = "end_date"),
  match_fun = list(`==`,`==`,`>=`,`<=`))

#adjusting taxa data
transect.dat <- transect.dat%>%
  rename("phylum" = "taxon_phylum_name", "class" = "taxon_class_name",
         "order" = "taxon_order_name", "family" = "taxon_family_name",
         "genus" = "taxon_genus_name", "species" = "taxon_species_name")

#giving transect data ecoregion and season

#some cleaning
transect.dat <-transect.dat%>%
  rename("state" = "state.x", "year" = "year.x")%>%
  mutate(state.y = NULL, year.y = NULL)

#creating jovial day then assigning season
dates <- as.Date(transect.dat$date, format = "%Y-%m-%d")
start_of_year <- as.Date(paste(transect.dat$year, "01", "01", sep = "-"), format = "%Y-%m-%d")
days_since_start <- as.numeric(dates - start_of_year)
transect.dat$nday <- days_since_start

transect.dat <- transect.dat%>%
  mutate(season = case_when(
    state %in% c("MA","CO") & nday < 195 ~ "early",
    state %in% c("MA","CO") & nday > 195 ~ "late",
    state %in% c("AK","AZ","KS","MN","OR") & nday < 155 ~ "early",
    state %in% c("AK","AZ","KS","MN","OR") & nday > 155 ~ "late",
    (state == "FL" & nday < 155) | (state == "FL" & nday > 310) ~ "early",
    state == "FL" & nday > 155 & nday < 310 ~ "late",
  ))

#assigning ecoregion based on state
transect.dat <- transect.dat%>%
  mutate(ecoregion = case_when ( state == "AK" ~ 'TAIG',state == "AZ" ~ 'DSW',
                                 state == "CO" ~ 'SROC',state == "FL" ~ 'SE',
                                 state == "KS" ~ 'PRAI',state == "MA" ~ 'NE',
                                 state == "MN" ~ 'GRTL',state == "OR" ~ 'PNW'))

#only genera that are found in the spore data
transect.dat.filtered <- transect.dat %>%
  filter(genus %in% macro.data.long$Genus)

length(unique(transect.dat.filtered$genus))

library(geosphere)

#location of each treatment
treatment_points <- spore.dat%>% filter(!is.na(Lat_dec)) %>%
  group_by(state,treatment) %>% summarise(long = mean(Long_dec), lat = mean(Lat_dec))

#distance between each treatment
temp <- distm(transect.dat.filtered[,c(5,4)], treatment_points[,c(3,4)], fun = distGeo)

#where is the nearest spore trap "treatment"
transect.dat.filtered$distance_m <- apply(temp, 1, FUN = min)

#making the otu table
transect_species <- transect.dat.filtered%>%
  mutate_if(is.character, list(~na_if(.,"")))%>%
  mutate(species = case_when(species == "" | is.na(species) ~ paste(genus, "sp", sep = " ")
                             ,TRUE ~ species))%>%
  mutate(season = case_when(season == "early" ~ "E", season == "late" ~ "L"))%>%
  mutate(ID = paste(ecoregion,year,season, sep = "_"))

#species column matching the spore data formatting
transect_species$species <- gsub(" ","_", transect_species$species)

transect_species%>% group_by(ecoregion)%>%summarise(n = n())%>% summarise(mean = mean(n))
transect_species%>% distinct(genus, ecoregion) %>% group_by(ecoregion)%>% summarise(n = n())%>% summarise(mean = mean(n))

write.csv(transect_species, "CLIMUSH_iNat_curated.csv")

```

```{r Proportion of Shared Genera & Breakpoint}

#making a key to pull multiple groups for the overlap
Overlap_key <- macro.data.long%>%
  mutate(key = paste(Genus_80, ecoregion, sep = "_")) %>%
  pull(key)

#getting the initial dataframe for the overlapping genera
iNat_prop <- transect.dat %>%
  mutate_if(is.character, list(~na_if(.,"")))%>%
  mutate(species = case_when(species == "" | is.na(species) ~ paste(genus, "sp", sep = "_")
                             ,TRUE ~ gsub(" ", "_", species)))%>%
  mutate(key = paste(genus, ecoregion, sep = "_"))%>%
  mutate(overlap = case_when(key %in% Overlap_key ~ "Overlap", TRUE ~ "None"))

spore_prop <- macro.data.long%>%
  filter(!is.na(ecoregion), !is.na(Genus_80))%>%
  filter(!str_detect(Genus_80,'sedis'))%>%
  mutate(key = paste(Genus_80, ecoregion, sep = "_")) %>%
  distinct(key)%>%
  mutate(overlap = case_when(key %in% iNat_prop$key ~ "Overlap", TRUE ~ "Spore Only"))%>%
  left_join(iNat_prop[,c(22,24,4,5)], by = 'key')

#distance between each treatment
temp <- distm(iNat_prop[,c(5,4)], treatment_points[,c(3,4)], fun = distGeo)

#where is the nearest spore trap "treatment"
iNat_prop$distance_m <- apply(temp, 1, FUN = min)

temp <- iNat_prop %>%
  group_by(ecoregion, genus) %>% summarise(distance_m = min(distance_m))

iNat_prop <-
  iNat_prop %>%
  right_join(temp, by = c("ecoregion","genus","distance_m"))


iNat_prop_summary <- iNat_prop %>%
  mutate(distance_bin = cut(
    distance_m, breaks = seq(0, 20000, by = 200),
    include.lowest = FALSE,right = FALSE)) %>%
  group_by(ecoregion, distance_bin) %>%
  # 3. Calculate overlap proportion
  summarise(
    total_n = n(),
    overlap_n = sum(overlap == "Overlap", na.rm = TRUE),
    .groups = "drop_last"
  ) %>%
  # Arrange bins numerically within each ecoregion
  #mutate(bin_start = as.numeric(str_extract(as.character(distance_bin), "\\d+(\\.\\d+)?"))) %>%
  arrange(ecoregion, distance_bin) %>%
  filter(!is.na(distance_bin))%>%
  # Compute cumulative totals and cumulative proportion
  mutate(
    cum_total = cumsum(total_n),
    cum_overlap = cumsum(overlap_n),
    cum_prop = cum_overlap / cum_total
  ) %>%
  ungroup()

breaks <- seq(0,20000,by = 200)
iNat_prop_summary <- iNat_prop_summary %>%
  mutate(bin_start = breaks[as.integer(distance_bin)])

#Fitting the Model to predict data
fit <- nls(cum_prop ~ a * exp(-b * bin_start)+c, 
           data = iNat_prop_summary, 
           start = list(a = 1, b = 0.0001, c = 0.1))
summary(fit)

#running prediction
iNat_prop_summary$fit <- predict(fit)

iNat_prop_summary <- iNat_prop_summary %>% filter(!is.na(ecoregion))

#finding the Breakpont
library(segmented)

lm_fit <- lm(log(cum_prop) ~ bin_start, data = iNat_prop_summary)
seg_fit <- segmented(lm_fit, seg.Z = ~bin_start, psi = list(bin_start = 1000))

summary(seg_fit)
print(seg_fit$psi)

#releveling ecoregions to keep colors consistent
library(viridis)

iNat_prop_summary%>%
  mutate(ecoregion = fct_relevel(ecoregion, c("SE","NE","PNW","SROC",
                                              "GRTL","TAIG","DSW","PRAI")))%>%
#graphing everything
ggplot(aes(x = bin_start/1000))+
  geom_point(aes(y = cum_prop, color = ecoregion), shape = 18, size = 2)+
  geom_line(aes(y = fit), color = 'black', size = 1)+
  geom_vline(xintercept = (seg_fit$psi[2])/1000, color = 'black', lty = 2, size = 0.5)+
  theme_classic()+
  scale_color_viridis(discrete = T)+
  labs(x = "Distance from Spore Trap (km)", y = "Proportion Genera Shared")


```

```{r Proportion Shared taxa one Ecoregion}

iNat_prop_single_summary <- iNat_prop %>%
  filter(ecoregion == "DSW")%>% #change for each ecoregion
  mutate(distance_bin = cut(
    distance_m, breaks = seq(0, 20000, by = 200),
    include.lowest = FALSE,right = FALSE)) %>%
  group_by(ecoregion, distance_bin) %>%
  # 3. Calculate overlap proportion
  summarise(
    total_n = n(),
    overlap_n = sum(overlap == "Overlap", na.rm = TRUE),
    .groups = "drop_last"
  ) %>%
  # Arrange bins numerically within each ecoregion
  #mutate(bin_start = as.numeric(str_extract(as.character(distance_bin), "\\d+(\\.\\d+)?"))) %>%
  arrange(ecoregion, distance_bin) %>%
  filter(!is.na(distance_bin))%>%
  # Compute cumulative totals and cumulative proportion
  mutate(
    cum_total = cumsum(total_n),
    cum_overlap = cumsum(overlap_n),
    cum_prop = cum_overlap / cum_total
  ) %>%
  ungroup()

breaks <- seq(0,20000,by = 200)
iNat_prop_single_summary <- iNat_prop_single_summary %>%
  mutate(bin_start = breaks[as.integer(distance_bin)])

library(minpack.lm)
#Fitting the Model to predict data
fit_sing <- nlsLM(
  cum_prop ~ a * exp(-b * bin_start) + c,
  data = iNat_prop_single_summary,
  start = list(a = 1, b = 0.001, c = 0.1)
)
summary(fit_sing)

#running prediction
iNat_prop_single_summary$fit <- predict(fit_sing)

iNat_prop_single_summary <- iNat_prop_single_summary %>% filter(!is.na(ecoregion))

#finding the Breakpont
library(segmented)

lm_fit_sing <- lm(log(cum_prop) ~ bin_start, data = iNat_prop_single_summary)
seg_fit_sing <- segmented(lm_fit_sing, seg.Z = ~bin_start, psi = list(bin_start = 1000))

#Switch this in for TAIG
#seg_fit_sing <- segmented(lm_fit_sing,seg.Z = ~bin_start,control = seg.control(display = TRUE, quant = TRUE))

summary(seg_fit_sing)

# Extract breakpoint and prepare label
bp_km <- seg_fit_sing$psi[2] / 1000

#releveling ecoregions to keep colors consistent
iNat_prop_single_summary%>%
#graphing everything
ggplot(aes(x = bin_start/1000))+
  geom_point(aes(y = cum_prop))+
  geom_line(aes(y = fit), color = 'black', size = 1)+
  geom_vline(xintercept = bp_km, color = 'black', lty = 2, size = 0.5)+
  geom_text(aes(x = bp_km, y = 0.75, 
      label = paste0(round(bp_km,2), " km")),
  inherit.aes = FALSE,
  hjust = -0.2, vjust = 0.1, size = 6)+
  theme_classic()+
  theme(plot.title = element_text(size = 20), axis.title.x = element_text(size = 15),
        axis.title.y = element_text(size = 15) )+
  scale_y_continuous(limits = c(0,0.8))+
  #change labs for each ecoregion
  labs(x = "Distance from Spore Trap (km)", y = "Proportion Genera Shared", title = "Desert Southwest") 

```
First step, I would like to make a species accumulation curve using the iNat data, sorting by the "nearest" observation to each spore trap. This is not habitat specific but IS state specific
```{r Species accumulation by ecoregion}

species.df <- transect_species %>%
  group_by(ecoregion, species) %>%
  summarise(min_distance_m = min(distance_m, na.rm = TRUE), .groups = "drop")

#comm_matrix <- species.df %>% select(ID, species)%>% mutate(presence = 1) %>% pivot_wider(names_from = species, values_from = presence, values_fill = 0)

#library(vegan)

#full species accumulation df
accum_df_full <- species.df %>%
  group_by(ecoregion) %>%
  arrange(min_distance_m, .by_group = TRUE) %>%
  mutate(
    cum_species = cumsum(!duplicated(species)),
    total_species = n_distinct(species),
    cum_prop = cum_species / total_species   # cumulative proportion
  ) %>%
  ungroup()

#fitting a line using Michaelis-Mentin (AIC selection)
nls_fit <- nls(cum_species ~ (a * min_distance_m) / (b + min_distance_m),
               data = accum_df_full,
               start = list(a = max(accum_df_full$cum_species), b = median(accum_df_full$min_distance_m)))


accum_df_full$fit <- predict(nls_fit)
coef_vals <- coef(nls_fit)
a <- coef_vals["a"]
b <- coef_vals["b"]
coef_vals <- as.data.frame(as.list(coef(nls_fit)))


#graphing all ecoregions together
accum_df_full%>%
mutate(ecoregion = fct_relevel(ecoregion, c("SE", "NE", "PNW","SROC","GRTL","TAIG","DSW","PRAI")))%>%
ggplot(aes(x = min_distance_m/1000, y = cum_species)) +
  geom_line(aes(color = ecoregion),size = 1, alpha = 0.7) +
  geom_line(aes(y = fit), color = "blue", size = 1.2)+
  geom_vline(data = coef_vals, aes(xintercept = b/1000), lty = 2)+
#both 'a' and 'b'
  #geom_text(data = coef_vals,
  #aes(x = b, y = 650, label = paste0("a = ", round(a)," species", "\nb = ", round(b)," meters")),
  #inherit.aes = FALSE,hjust = -0.1, vjust = 0, size = 6)+
#doing 'b' only to match the breakpoints
  geom_text(data = coef_vals,
  aes(x = b/1000, y = 650, label = paste0(round(b/1000,2)," km")),
  inherit.aes = FALSE,hjust = -0.1, vjust = 0, size = 6)+
  scale_color_viridis(discrete = T)+
  labs(x = "Distance from trap (km)", y = "Cumulative Number of Species",
       title = "", color = "Ecoregion") +
  theme_classic()

```

```{r SAC with range estimations by ecoregion}
library(ggplot2)
library(dplyr)

#subsetting to run breakpoints below - change for each ecoregion
accum_df <- accum_df_full %>% filter(ecoregion == "PNW")

# Fit Michaelis-Menten model
nls_fit <- nls(
  cum_species ~ (a * min_distance_m) / (b + min_distance_m),
  data = accum_df,
  start = list(
    a = max(accum_df$cum_species, na.rm = TRUE),
    b = median(accum_df$min_distance_m, na.rm = TRUE)
  )
)

# Get fitted values and coefficients
accum_df$fit <- predict(nls_fit)
coef_vals <- coef(nls_fit)
a <- coef_vals["a"]
b <- coef_vals["b"]
coef_vals <- as.data.frame(as.list(coef(nls_fit)))

# Build a smooth curve across the observed x range
fit_df <- data.frame(
  min_distance_m = seq(min(accum_df$min_distance_m),
                       max(accum_df$min_distance_m),
                       length.out = 200)
) %>%
  mutate(fit = (a * min_distance_m) / (b + min_distance_m))

#parametric bootstrap for confidence intervals
boot_preds <- replicate(500, {
  new_a <- rnorm(1, mean = a, sd = summary(nls_fit)$coefficients["a","Std. Error"])
  new_b <- rnorm(1, mean = b, sd = summary(nls_fit)$coefficients["b","Std. Error"])
  (new_a * fit_df$min_distance_m) / (new_b + fit_df$min_distance_m)
})

fit_df$lower <- apply(boot_preds, 1, quantile, 0.025)
fit_df$upper <- apply(boot_preds, 1, quantile, 0.975)

# Plot
fit_df$cum_species <- row_number(fit_df$min_distance_m)

#putting location on the coefficients
coef_vals <- coef_vals %>%
  mutate(x = min(accum_df$min_distance_m, na.rm = TRUE) * 1.1,
         y = max(accum_df$cum_species, na.rm = TRUE) * 0.9,
         label = paste0("a = ", round(a, 2), 
                        ", b = ", round(b, 2)))



ggplot(accum_df, aes(x = min_distance_m/1000, y = cum_species)) +
  geom_line(color = 'black', lwd = .8)+
  geom_line(data = fit_df, aes(y = fit), color = "blue", linewidth = 1) +
  geom_vline(data = coef_vals, aes(xintercept = b/1000), lty = 2)+
  geom_ribbon(data = fit_df, aes(ymin = lower, ymax = upper), 
              fill = "blue", alpha = 0.2) +
  theme_classic() +
#just 'b' to keep it cleaner
  geom_label(data = coef_vals, 
            aes(x = 2, y = y, label = paste0(round(b/1000, 2), " km")),
            hjust = -0.1,
#creating a white box behind the label to ensure the line never intersects the 'b' stat
            fill = "white", linewidth = 0, label.padding = unit(0.2, 'lines'), size = 10)+
  labs(x = "Distance from Spore Trap (m)",
       y = "Cumulative Species Richness",
       title = "Taiga")+ #change titles for each ecoregion
  guides(color = 'none')+
  scale_x_continuous(limits = c(0,35))+ #fixing axes so any extreme 'b' doesnt change the graph
  theme(plot.title = element_text(size = 30))

```
